# Exploring Red Wine Chemical Properties and Quality
## Sean Stevenson
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

WINE_COLOR = "#722F37"
COMP_COLOR = "#2F726A"

library(ggplot2)
library(gridExtra)
```

I will be examining a data set of 1599 red wines consisting of 11 physiochemical
properties as input variables, as well as a rating for each wine 
as an output variable. Rating is defined as the median
score given by at least three different wine experts on a scale of 0 (worst) to
10 (best).


After loading the data and doing some minor cleaning, the structure of the data
set can be seen below:

```{r echo=FALSE, Load_the_Data}
# Load the Data

wine = read.csv('wineQualityReds.csv')
#remove redundant indexing
wine$X <- NULL

#rename quality to rating; quality will be used for something else later
wine$rating <- wine$quality
wine$quality <- NULL

str(wine)
```

Below is a description of all variables, taken from the [data set readme]
(https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt):

> 1. fixed acidity: most acids involved with wine are fixed or nonvolatile (do 
not evaporate readily)
> 2. volatile acidity: the amount of acetic acid in wine, which at too high of 
levels can lead to an unpleasant, vinegar taste
> 3. citric acid: found in small quantities, citric acid can add 'freshness' and
flavor to wines
> 4. residual sugar: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater than 
45 grams/liter are considered sweet
> 5. chlorides: the amount of salt in the wine
> 6. free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
> 7. total sulfur dioxide: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
> 8. density: the density of wine is close to that of water depending on the 
percent alcohol and sugar content
> 9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic)
to 14 (very basic); most wines are between 3-4 on the pH scale
> 10. sulphates: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, which acts as an antimicrobial and antioxidant
> 11. alcohol: the percent alcohol content of the wine
> 12. rating: (score between 0 and 10)

# Univariate Plots Section


```{r echo=FALSE, Rating_Plot}
ggplot(wine, aes(rating)) +
  geom_histogram(binwidth = 1, fill = WINE_COLOR, colour="black") + 
  scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 10)) + 
  scale_y_continuous(breaks = seq(0, 700, 100)) +
  ggtitle("Distribution of Wine Ratings")


```

Most wines have a rating of either 5 or 6 and all wines fall within the 3-8
range. The low variance of ratings is somewhat unsurprising
given each rating is actually the median rating of at least 3 different experts.
Because of the relative sparsity of data outside the 5-6 range, I am going to
create a new factor variable called 'quality' to classify each wine as either
'good' if it has an above average rating (>= 6) or 'mediocre' if it has a 
below average rating (<= 5). This simplification will make it easier to see any
patterns that might characterize good or bad wines.

```{r echo=FALSE, Alcohol_Plot}

# create new factor variable 'quality', good if rating >= 6 and mediocre if not
wine$quality <- as.factor(ifelse(wine$rating >= 6, 'good', 'mediocre'))


ggplot(wine, aes(alcohol)) + 
  geom_histogram(binwidth = .5, fill=WINE_COLOR, colour="black") + 
  scale_x_continuous(breaks = seq(0, 20, .5)) + 
  xlab("alcohol (% by volume)") + 
  ggtitle("Distribution of Alcohol Content")

ggplot(wine, aes(alcohol)) + 
  geom_freqpoly(binwidth = .5, aes(color = quality)) + 
  scale_x_continuous(breaks = seq(0, 20, .5)) +
  xlab("alcohol (% by volume)") + 
  ggtitle("Distribution of Alcohol Content")

summary(wine$alcohol)
```

```{r echo=FALSE, Acidity_Plots}

# acidity plots
citric <- ggplot(wine, aes(citric.acid)) + 
  geom_freqpoly(binwidth = .02) +
  xlab("citric acid (g/dm^3)")
volatile <- ggplot(wine, aes(volatile.acidity)) + geom_freqpoly()
fixed <- ggplot(wine, aes(fixed.acidity)) + geom_freqpoly()
pH <- ggplot(wine, aes(pH)) + geom_freqpoly(binwidth = .05) +
  scale_x_continuous(breaks = seq(0, 14, .2))
grid.arrange(citric, volatile, fixed, pH, ncol = 2, top = 'Distributions of Acids')
```

```{r echo=FALSE, Density_Plots}

ggplot(wine, aes(density)) + geom_freqpoly(binwidth = .001)
```

```{r echo=FALSE, Sugar_Plot}
ggplot(wine, aes(residual.sugar)) + geom_freqpoly()

ggplot(wine, aes(chlorides)) + geom_histogram()
```

```{r echo=FALSE, SO2_Plot}
# so2 plots 
free_so2 <- ggplot(wine, aes(free.sulfur.dioxide)) + geom_freqpoly()
total_so2 <- ggplot(wine, aes(total.sulfur.dioxide)) + geom_freqpoly()
sulphates <- ggplot(wine, aes(sulphates)) + geom_freqpoly()
grid.arrange(free_so2, total_so2, sulphates)

free_so2 <- ggplot(wine, aes(free.sulfur.dioxide)) + geom_freqpoly(aes(color = quality))
total_so2 <- ggplot(wine, aes(total.sulfur.dioxide)) + geom_freqpoly(aes(color = quality))
sulphates <- ggplot(wine, aes(sulphates)) + geom_freqpoly(aes(color = quality))
grid.arrange(free_so2, total_so2, sulphates)

citric <- ggplot(wine, aes(citric.acid)) + 
  geom_freqpoly(binwidth = .02, aes(color = quality)) +
  xlab("citric acid (g/dm^3)")
volatile <- ggplot(wine, aes(volatile.acidity)) + geom_freqpoly(aes(color = quality))
fixed <- ggplot(wine, aes(fixed.acidity)) + geom_freqpoly(aes(color = quality))
pH <- ggplot(wine, aes(pH)) + geom_freqpoly(binwidth = .05, aes(color = quality)) +
  scale_x_continuous(breaks = seq(0, 14, .2))
grid.arrange(citric, volatile, fixed, pH, ncol = 2, top = 'Distributions of Acids')


```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?



### What is/are the main feature(s) of interest in your dataset?

The primary feature of interest in the data set is the wine ratings. This is the
feature I am interested in predicting

### What other features in the dataset do you think will help support your \
All of the physiochemical properties will need to be investigated to see their
relationship with wine ratings/quality.

### Did you create any new variables from existing variables in the dataset?

A new factor variable 'quality' was created classifying wines as either good or
mediocre depending on their rating.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I removed the indexing variable and renamed quality to rating to use quality
for a variable of my own.


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}
# TODO: see how various acidic compounds and related

# TODO: see how everything effects rating/quality

ggplot(wine, aes(x = residual.sugar, y = alcohol)) + geom_point(aes(color = quality))

ggplot(wine, aes(x = alcohol, y = pH)) + geom_point(aes(color = quality))

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!